---
title: "Adam Lyon UTSW Sample Processing and Figure Generation"
author: "Samuel Cooke"
date: "2025-04-17"
output: html_document
---

### Single Cell Seurat Data Analysis
## Load Necessary Packages
```{r import, warning=FALSE,message=FALSE}
library(reshape2)
library(tidyverse)
library(MASS)
library(Matrix)
library(MatrixGenerics)
library(edgeR)
library(mltools)
library(pheatmap)
library(data.table)
library(Seurat)
library(SeuratData)
library(knitr)
library(dplyr)
library(patchwork)
```
                      
## Check Working Directory
```{r Working Directory}
print(getwd())
```

## Load the data
```{r load_data, warning=FALSE,message=FALSE}

filtered_h5 <- "./data/Filtered_Features/aggr_5removed_filtered_feature_bc_matrix.h5"

counts <- Read10X_h5(filtered_h5)

UTSW_seurat <- CreateSeuratObject(counts = counts, project = "UTSW Aggregated Samples")

```

## Extract sample information into meta.data
```{r, sample names}

# 1. Define mapping from suffix number → sample name
sample_levels <- c("3a","3b","4a","4b","7a","7b")

# 2. Grab your cell barcodes
barcodes <- colnames(UTSW_seurat)

# 3. Extract the numeric suffix after the dash
#    e.g. "AACAACCCAGCTATAC‑1" → "1", coerced to integer 1
suffix_nums <- as.integer(sub(".*-(\\d+)$", "\\1", barcodes))

# 4. Map those numbers to your sample names
sample_names <- sample_levels[suffix_nums]

# 5. Add to meta.data (as a factor, keeping the proper order)
UTSW_seurat$sample <- factor(sample_names, levels = sample_levels)

# 6. (Optional) Check it
before_filtering <- table(UTSW_seurat$sample)
before_filtering
```

## Sample Filtering
```{r, filter}

umi_cutoffs <- list(
  "3a" = 27,
  "3b" = 27,
  "4a" = 60,
  "4b" = 70,
  "7a" = 65,
  "7b" = 35
)

# 1. compute per‐cell UMI counts (if you haven’t already)
umi_counts <- Matrix::colSums(
  GetAssayData(UTSW_seurat, assay = "RNA", slot = "counts")
)

# 2. grab your cell‐by‐cell sample assignments
samples     <- UTSW_seurat$sample

# 3. build the “keep?” logical vector using your named umi_cutoffs
keep <- umi_counts >= umi_cutoffs[samples]

# 4. extract the cell barcodes that pass
cells_to_keep <- names(keep)[keep]

# 5. subset in one go
UTSW_filtered <- subset(UTSW_seurat, cells = cells_to_keep)
after_filtering <- table(UTSW_filtered$sample)

```

## Sample Filtering Report
```{r, filtering table}

# 1. build the two tables
before_filtering <- table(UTSW_seurat$sample)
after_filtering  <- table(UTSW_filtered$sample)

# 2. coerce the “before” table to a data.frame
df <- as.data.frame(before_filtering)
colnames(df) <- c("sample", "before_filtering")

# 3. add the “after” counts by looking them up by sample name
df$after_filtering <- as.integer(after_filtering[ as.character(df$sample) ])

# 4. compute percent filtered out
df$percent_filtered_out <- 
  (df$before_filtering - df$after_filtering) / df$before_filtering * 100

# 5. (optional) round to one decimal place
df$percent_filtered_out <- round(df$percent_filtered_out, 1)

# 6. view
print(df)

```

## Visualize data Before Filtering
```{r, visualize QC}

# 1. Add complexity metric to your Seurat object
UTSW_seurat$log10GenesPerUMI <- 
  log10(UTSW_seurat$nFeature_RNA) / log10(UTSW_seurat$nCount_RNA)

# 2. Extract & rename metadata columns
metadata <- UTSW_seurat@meta.data
metadata$cells <- rownames(metadata)

metadata <- metadata %>%
  dplyr::rename(seq_folder = orig.ident,
                nUMI = nCount_RNA,
                nGene = nFeature_RNA)

# 3. QC plots

## UMI Count Distribution per Sample
ggplot(metadata, aes(x = nUMI, color = sample, fill = sample)) + 
  geom_density(alpha = 0.2) + 
  scale_x_log10() + 
  theme_classic() +
  ylab("Cell density") +
  ggtitle("UMI Count Distribution per Sample") +
  geom_vline(xintercept = 1000, linetype = "dashed") +
  geom_vline(xintercept =   25, linetype = "dashed")

## Gene Count Distribution per Sample
ggplot(metadata, aes(x = nGene, color = sample, fill = sample)) + 
  geom_density(alpha = 0.2) + 
  scale_x_log10() + 
  theme_classic() +
  ggtitle("Gene Count Distribution per Sample") +
  geom_vline(xintercept =   6, linetype = "dashed") +
  geom_vline(xintercept = 100, linetype = "dashed")

## Boxplot of Genes per Cell (log10 scaled)
ggplot(metadata, aes(x = sample, y = log10(nGene), fill = sample)) + 
  geom_boxplot() + 
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Genes per Cell (log10 scaled)")

## Correlation nUMI vs nGene by Sample (faceted)
ggplot(metadata, aes(x = nUMI, y = nGene, color = sample)) + 
  geom_point(alpha = 0.4, size = 0.5) + 
  stat_smooth(method = "lm", se = FALSE, color = "black") +
  scale_x_log10() + 
  scale_y_log10() + 
  theme_classic() +
  geom_vline(xintercept = 500, linetype = "dashed") +
  geom_hline(yintercept = 250, linetype = "dashed") +
  facet_wrap(~ sample)

## nUMI vs nGene Across All Samples (overlay)
ggplot(metadata, aes(x = nUMI, y = nGene, color = sample)) + 
  geom_point(alpha = 0.4, size = 0.5) + 
  stat_smooth(method = "lm", se = FALSE, color = "black") +
  scale_x_log10() + 
  scale_y_log10() + 
  theme_classic() +
  ggtitle("nUMI vs nGene Across All Samples (Overlayed)") +
  theme(plot.title = element_text(hjust = 0.5))

## Gene Complexity per Sample
ggplot(metadata, aes(x = log10GenesPerUMI, color = sample, fill = sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  geom_vline(xintercept = 0.8, linetype = "dashed") +
  ggtitle("Gene Complexity per Sample")

## Histogram of Gene Complexity (faceted)
ggplot(metadata, aes(x = log10GenesPerUMI, fill = sample)) +
  geom_histogram(binwidth = 0.02, color = "black", alpha = 0.6) +
  theme_classic() +
  facet_wrap(~ sample, scales = "free_y") +
  scale_x_continuous(limits = c(0.8, NA)) +
  labs(
    title = "Histogram of log10(Genes) / log10(UMIs)", 
    x     = "log10(Genes) / log10(UMIs)", 
    y     = "Cell Count"
  )

## Violin Plot of Genes per Cell (log10 scaled)
ggplot(metadata, aes(x = sample, y = log10(nGene), fill = sample)) + 
  geom_violin(trim = FALSE, scale = "width") +
  theme_classic() +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1),
    plot.title   = element_text(hjust = 0.5, face = "bold")
  ) +
  ggtitle("Genes per Cell (log10 scaled)") +
  ylab("log10(nGene)")

```

## Visualize Data After Filtering
```{r, visualize QC after filtering}

# 1. Add complexity metric to your Seurat object
UTSW_filtered$log10GenesPerUMI <- 
  log10(UTSW_filtered$nFeature_RNA) / log10(UTSW_filtered$nCount_RNA)

# 2. Extract & rename metadata columns
metadata_filtered <- UTSW_filtered@meta.data
metadata_filtered$cells <- rownames(metadata_filtered)

metadata_filtered <- metadata_filtered %>%
  dplyr::rename(seq_folder = orig.ident,
                nUMI = nCount_RNA,
                nGene = nFeature_RNA)

# 3. QC plots

## UMI Count Distribution per Sample
ggplot(metadata_filtered, aes(x = nUMI, color = sample, fill = sample)) + 
  geom_density(alpha = 0.2) + 
  scale_x_log10() + 
  theme_classic() +
  ylab("Cell density") +
  ggtitle("UMI Count Distribution per Sample") +
  geom_vline(xintercept = 1000, linetype = "dashed") +
  geom_vline(xintercept =   25, linetype = "dashed")

## Gene Count Distribution per Sample
ggplot(metadata_filtered, aes(x = nGene, color = sample, fill = sample)) + 
  geom_density(alpha = 0.2) + 
  scale_x_log10() + 
  theme_classic() +
  ggtitle("Gene Count Distribution per Sample") +
  geom_vline(xintercept =   6, linetype = "dashed") +
  geom_vline(xintercept = 100, linetype = "dashed")

## Boxplot of Genes per Cell (log10 scaled)
ggplot(metadata_filtered, aes(x = sample, y = log10(nGene), fill = sample)) + 
  geom_boxplot() + 
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Genes per Cell (log10 scaled)")

## Correlation nUMI vs nGene by Sample (faceted)
ggplot(metadata_filtered, aes(x = nUMI, y = nGene, color = sample)) + 
  geom_point(alpha = 0.4, size = 0.5) + 
  stat_smooth(method = "lm", se = FALSE, color = "black") +
  scale_x_log10() + 
  scale_y_log10() + 
  theme_classic() +
  geom_vline(xintercept = 500, linetype = "dashed") +
  geom_hline(yintercept = 250, linetype = "dashed") +
  facet_wrap(~ sample)

## nUMI vs nGene Across All Samples (overlay)
ggplot(metadata_filtered, aes(x = nUMI, y = nGene, color = sample)) + 
  geom_point(alpha = 0.4, size = 0.5) + 
  stat_smooth(method = "lm", se = FALSE, color = "black") +
  scale_x_log10() + 
  scale_y_log10() + 
  theme_classic() +
  ggtitle("nUMI vs nGene Across All Samples (Overlayed)") +
  theme(plot.title = element_text(hjust = 0.5))

## Gene Complexity per Sample
ggplot(metadata_filtered, aes(x = log10GenesPerUMI, color = sample, fill = sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  geom_vline(xintercept = 0.8, linetype = "dashed") +
  ggtitle("Gene Complexity per Sample")

## Histogram of Gene Complexity (faceted)
ggplot(metadata_filtered, aes(x = log10GenesPerUMI, fill = sample)) +
  geom_histogram(binwidth = 0.02, color = "black", alpha = 0.6) +
  theme_classic() +
  facet_wrap(~ sample, scales = "free_y") +
  scale_x_continuous(limits = c(0.8, NA)) +
  labs(
    title = "Histogram of log10(Genes) / log10(UMIs)", 
    x     = "log10(Genes) / log10(UMIs)", 
    y     = "Cell Count"
  )

## Violin Plot of Genes per Cell (log10 scaled)
ggplot(metadata_filtered, aes(x = sample, y = log10(nGene), fill = sample)) + 
  geom_violin(trim = FALSE, scale = "width") +
  theme_classic() +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1),
    plot.title   = element_text(hjust = 0.5, face = "bold")
  ) +
  ggtitle("Genes per Cell (log10 scaled)") +
  ylab("log10(nGene)")

```

## Collapse Samples
```{r, collapse samples}

# ──────────────────────────────────────────────────────────────
# Collapse the sample labels: 3a+3b → 3, 4a+4b → 4, 7a+7b → 7
# ──────────────────────────────────────────────────────────────

# Collapse 3a→3, 3b→3, etc. by removing the final letter
UTSW_filtered$sample_collapsed <- factor(
  sub("[ab]$", "",               # regex: drop last a or b
      as.character(UTSW_filtered$sample)),
  levels = c("3","4","7")        # keep desired order
)

# sanity-check
table(original  = UTSW_filtered$sample,
      collapsed = UTSW_filtered$sample_collapsed)

UTSW_filtered$sample <- UTSW_filtered$sample_collapsed

UTSW_filtered$sample_collapsed <- NULL

```

## Convert 3 to Parent
```{r, convert to parent}

# ── Change "3" → "parent" in metadata$sample ──
# 1) If the column is a factor, convert to character so we can alter the values
UTSW_filtered$sample <- as.character(UTSW_filtered$sample)

# 2) Replace every "3" with "parent"
UTSW_filtered$sample[UTSW_filtered$sample == "3"] <- "parent"

# 3) (Optional) Convert back to a factor and set the level order you want
UTSW_filtered$sample <- factor(UTSW_filtered$sample,
                             levels = c("parent", "4", "7"))

```

## Probe collapse based on max counts
```{r, Seurat_Probe_Collapse, message=FALSE, warning=FALSE}

# Ensure RNA is active assay
DefaultAssay(UTSW_filtered) <- "RNA"

# 1. Get the raw (counts) assay data
counts_mat <- GetAssayData(object = UTSW_filtered, slot = "counts")

# 2. Sum counts per feature (rowSums is efficient for dgCMatrix objects)
feature_sums <- Matrix::rowSums(counts_mat)

# 3. Create a data frame of features and their summed counts

feature_counts_df <- data.frame(
  Feature       = names(feature_sums), 
  Summed_Counts = feature_sums
)

# 4. Split the Feature Names into Gene and Probe
## 'Feature' is something like "CD79A-1".
## strsplit(...) will return a list of character vectors, one per row.

split_list <- strsplit(feature_counts_df$Feature, split = "-")
feature_counts_df$Gene  <- sapply(split_list, `[`, 1)
feature_counts_df$Probe <- sapply(split_list, `[`, 2)

# 5. Reorder the matrix to Feature, Gene, Probe, Summed_Counts
feature_counts_df <- feature_counts_df[, c("Feature", "Gene", "Probe", "Summed_Counts")]

# 6. Extract each probe with the max Summed_Counts compared

max_probe_df <- feature_counts_df %>%
  group_by(Gene) %>%
  slice_max(
    order_by  = Summed_Counts,
    n         = 1,
    with_ties = FALSE
  ) %>%
  ungroup()

# 7. Extract all the probe names to keep
## The 'Feature' column here names the rows (probes) to keep for each Gene
keep_features <- max_probe_df$Feature

# 8. Makes a copy of the original counts matrix extracted from the seurat object
new_mat <- counts_mat

# 9. Zeros out all probes not in the keep_features list
## For all rows not in 'keep_features', set them to zero
not_keep <- setdiff(rownames(new_mat), keep_features)
new_mat[not_keep, ] <- 0

# 10. Create a new Assay object
new_assay <- CreateAssayObject(counts = new_mat)

# 11. Add this assay to the existing Seurat object under a chosen name (e.g., "ProbeCollapseAssay")
UTSW_filtered[["ProbeCollapseAssay"]] <- new_assay

```

## Seurat Standard Processing
```{r, seurat processing}

DefaultAssay(UTSW_filtered) <- "ProbeCollapseAssay"

# Standard Seurat Processing
UTSW_filtered <- NormalizeData(UTSW_filtered)
UTSW_filtered <- FindVariableFeatures(UTSW_filtered)
UTSW_filtered <- ScaleData(UTSW_filtered)
UTSW_filtered <- RunPCA(UTSW_filtered, verbose = FALSE)
UTSW_filtered <- FindNeighbors(UTSW_filtered, dims = 1:7)
UTSW_filtered <- FindClusters(UTSW_filtered, resolution = 0.7, verbose = FALSE)
UTSW_filtered <- RunUMAP(UTSW_filtered, dims = 1:7)
UTSW_filtered_UMAP <- DimPlot(UTSW_filtered, label = TRUE)
UTSW_filtered_UMAP_Sample <- DimPlot(UTSW_filtered, label = TRUE, group.by = "sample")
# Display UMAP
UTSW_filtered_UMAP
UTSW_filtered_UMAP_Sample


# export figure
ggsave(
  filename = "figures/UTSW_UMAP.pdf",
  plot     = UTSW_filtered_UMAP,      # the object you created
  width    = 6,                # ← 20 inches wide
  height   = 4,                 # ← 25 inches tall
  units    = "in",              # "in", "cm", or "mm"
  device   = cairo_pdf          # good font embedding on most systems
)

ggsave(
  filename = "figures/UTSW_UMAP_Sample.pdf",
  plot     = UTSW_filtered_UMAP_Sample,      # the object you created
  width    = 6,                # ← 20 inches wide
  height   = 4,                 # ← 25 inches tall
  units    = "in",              # "in", "cm", or "mm"
  device   = cairo_pdf          # good font embedding on most systems
)

```

## Split UMAP
```{r, split umap, fig.width = 20, fig.height = 6}

UTSW_UMAP_Split<- DimPlot(
  UTSW_filtered,
  reduction = "umap",
  # colour however you like:
  group.by = "seurat_clusters",   # or "sample", "celltype", etc.
  # ← this is the key line:
  split.by = "sample",            # one facet per sample, only its cells are shown
  ncol     = 3,                   # arrange the tiles 3-across (tweak as desired)
  pt.size  = 0.2                  # keep the dots small when faceting
) + NoLegend()

UTSW_UMAP_Split

ggsave(
  filename = "figures/UTSW_UMAP_Split.pdf",
  plot     = UTSW_UMAP_Split,      # the object you created
  width    = 20,                # ← 20 inches wide
  height   = 6,                 # ← 25 inches tall
  units    = "in",              # "in", "cm", or "mm"
  device   = cairo_pdf          # good font embedding on most systems
)

```

## Gene Distribution Graphs
```{r, dstribution graphs, fig.width = 12, fig.height = 3.5, fig.keep = 'all'}

# ---------------------------------------------------------------
#  1) Make list containing genes of interest
# ---------------------------------------------------------------

genes <- c(
  "hipA","hipB","gatA","glvC","selB","rpoZ","ftsH","fimE","fimA","lpxC",
  "kdtA","gyrA","rpoB","icd","fusA","trkH","phoQ","phoP","prs","cpxA",
  "relA","spoT","lon","sulA","yqgE","ompC","acrA","pgaC","ygfB","sbmA",
  "rpoH","sucA","priA","dnaC","dnaT","cysK","mdtK","yhaM","rmf","oxyR",
  "rpoS","dksA","gsk","nadC","purF","ahpF","katG","poxB","acs","lsrA",
  "ptsG","aceK","rplJ","feaB","puuA","elaD","recF","paaG","pssL","arcA",
  "arpA","tas","yhiI","arcB","hlyE","xerC","osmY","yaaA","yacH","glpF",
  "uvrC","pgsA","baeS","tufB","bluF","pitA","corA","yicC","eutH","relB",
  "yfbS","cstA","abgT","glnH","yjbB","nadK","waaF","cysS","marR","pheS",
  "rbsA","glcB","yqeF","yieK","thiP","ptsI","fadB","eutQ","torC","dnaQ",
  "topA","rpoD","nikA","aroG","mutL","rihB","serW","speE","clsB","rnb",
  "mscM","cyaA","thpR","yagN","mltB","ydcL","ydgI","glpT","yebK","yfjH",
  "tyrR","hns","tdk","ppa","ytfQ","ampH","narU","yddK","appY","yjcO",
  "gltP","ompT","ftp","yffN","yffO","mrcB","mcrB","symE","yohP","dusC",
  "rssB","galU","gadF","mdtE","rnpB","yhaC","manZ","yobD","yhfG","ppiA",
  "adiY","adiA","iscR","trmJ","mgtL","mgtA"
)

# ---------------------------------------------------------------
#  2) Identify the retained probe (gene-#) for each gene of interest
# ---------------------------------------------------------------

# Make sure the collapsed assay is active
DefaultAssay(UTSW_filtered) <- "ProbeCollapseAssay"

# Pull the counts matrix from the collapsed assay
collapsed_mat <- GetAssayData(UTSW_filtered, slot = "counts")

# Helper: given a gene root (e.g. "hipA"), return the probe that still has counts
kept_probe <- function(gene_name, mat) {
  
  # Rows that start with "gene-" (all probes for that gene)
  probe_rows <- grep(paste0("^", gene_name, "-"), rownames(mat), value = TRUE)
  
  if (length(probe_rows) == 0) return(NA_character_)   # gene absent
  
  # Sum counts across all cells for every probe
  totals <- Matrix::rowSums(mat[probe_rows, , drop = FALSE])
  
  # After collapsing, exactly one probe should have non-zero counts
  nz <- probe_rows[totals > 0]
  
  if (length(nz) == 0) {
    NA_character_                         # none retained
  } else if (length(nz) == 1) {
    nz                                    # one retained
  } else {
    # Rare tie: keep the probe with the *largest* total counts
    nz[which.max(totals[probe_rows %in% nz])]
  }
}

# Build a tidy data frame: Gene  |  SelectedProbe  |  ProbeID
kept_probe_df <- data.frame(
  Gene          = genes,
  SelectedProbe = vapply(genes, kept_probe, character(1), mat = collapsed_mat),
  stringsAsFactors = FALSE
)

# Extract just the numeric part after the dash
kept_probe_df$ProbeID <- sub(".*-(\\d+)$", "\\1", kept_probe_df$SelectedProbe)

# ---------------------------------------------------------------
# 3) Plot genes of interest distribution on UMAP
# ---------------------------------------------------------------

# probes that were actually retained in the collapsed assay
genes_probes <- as.character(na.omit(kept_probe_df$SelectedProbe))

n_samples <- length(levels(UTSW_filtered$sample))   # panels per figure

# pre-allocate named list
plot_list <- vector("list", length(genes_probes))
names(plot_list) <- genes_probes

for (i in seq_along(genes_probes)) {
  
  gene_id <- genes_probes[i]
  
  p <- FeaturePlot(
        object    = UTSW_filtered,
        features  = gene_id,          # one probe at a time
        reduction = "umap",
        split.by  = "sample",
        ncol      = n_samples,        # all sample tiles in one row
        pt.size   = 0.25,
        order     = TRUE,
        raster    = FALSE
      ) &
      plot_annotation(title = gene_id) &
      theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  
  plot_list[[i]] <- p          # save into list, no printing
}

# Display all plots
#plot_list

```

## Save Distribution Plots
```{r, dis plots save}

# 1) Make sure the output folder exists
output_dir <- "figures/gene_distribution"
if (!dir.exists(output_dir)) dir.create(output_dir)

# 2) Iterate over the named list and save each plot
for (probe_id in names(plot_list)) {
  
  p <- plot_list[[probe_id]]
  
  # Sanitise the filename in case probe_id has characters like "/"
  safe_name <- gsub("[^A-Za-z0-9_\\-]", "_", probe_id)
  file_out  <- file.path(output_dir, paste0(safe_name, ".pdf"))
  
  ggsave(
    filename = file_out,
    plot     = p,
    device   = "pdf",
    dpi      = 300,
    width    = 20,   # adjust if you want larger/smaller files
    height   = 6,
    units    = "in"
  )
}

cat("Saved", length(plot_list), "plots to", output_dir, "\n")

```
## Annotated Charts
```{r, annotated distribution DGE}

## ──────────────────────────────────────────────────────────────
## Compute DGE for Parent, 4, and 7 using shared parameters
## Restrict to `genes_probes` only
## ──────────────────────────────────────────────────────────────

# Parameters
logfc_threshold <- 0
min_pct         <- 0
only_pos        <- FALSE

# Samples to compare (against all others)
sample_ids <- c("parent", "4", "7")
names(sample_ids) <- c("Parent", "Sample 4", "Sample 7")  # pretty names

# Compute DGE and store in a named list
dge_list <- lapply(sample_ids, function(smpl) {
  FindMarkers(
    UTSW_filtered,
    ident.1         = smpl,
    group.by        = "sample",
    assay           = "ProbeCollapseAssay",
    features        = genes_probes,
    logfc.threshold = logfc_threshold,
    min.pct         = min_pct,
    only.pos        = only_pos
  )
})

# Add names to the list
names(dge_list) <- names(sample_ids)



```

## Distribution Plots with DGE Data Chart
```{r, Annotated Distribution Plots}

## ──────────────────────────────────────────────────────────────
## Combine FeaturePlot + gene regulation summary table
## ──────────────────────────────────────────────────────────────

library(gridExtra)
library(grid)
library(patchwork)

# Helper: Generate regulation table per gene
get_gene_status_table <- function(probe_id, dge_list) {
  
  # Build one row per sample
  res <- lapply(names(dge_list), function(smpl_name) {
    dge <- dge_list[[smpl_name]]
    
    if (!probe_id %in% rownames(dge)) {
      return(data.frame(
        Sample    = smpl_name,
        Direction = "NA",
        Log2FC    = "NA",
        p_val_adj = "NA",
        stringsAsFactors = FALSE
      ))
    }
    
    logfc <- dge[probe_id, "avg_log2FC"]
    padj  <- dge[probe_id, "p_val_adj"]
    dir   <- ifelse(logfc > 0, "Up", "Down")
    
    data.frame(
      Sample    = smpl_name,
      Direction = dir,
      Log2FC    = sprintf("%.2f", logfc),
      p_val_adj = formatC(padj, format = "e", digits = 2),
      stringsAsFactors = FALSE
    )
  })
  
  # Combine rows into a data frame
  df <- do.call(rbind, res)
  
  # Create table grob for plotting
  tableGrob(df, rows = NULL, theme = ttheme_default(base_size = 8))
}

# Initialize Plot List
plot_list <- vector("list", length(genes_probes))
names(plot_list) <- genes_probes

# Combine UMAP plot with annotation table
for (i in seq_along(genes_probes)) {
  
  gene_id <- genes_probes[i]
  
  # UMAP split-by-sample
  p <- FeaturePlot(
    object    = UTSW_filtered,
    features  = gene_id,
    reduction = "umap",
    split.by  = "sample",
    ncol      = length(levels(UTSW_filtered$sample)),
    pt.size   = 0.25,
    order     = TRUE,
    raster    = FALSE
  ) &
  plot_annotation(title = gene_id) &
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  
  # Table of Up/Down + log2FC
  table_grob <- get_gene_status_table(gene_id, dge_list)
  
  # Combine vertically: UMAP on top, table on bottom
  combined <- p / wrap_elements(table_grob) + plot_layout(heights = c(4, 1))
  
  # Store
  plot_list[[i]] <- combined
}

```

## Save Distribution Plots 2
```{r, dis plots save 2}

# 1) Make sure the output folder exists
output_dir <- "figures/gene_distribution_charts"
if (!dir.exists(output_dir)) dir.create(output_dir)

# 2) Iterate over the named list and save each plot
for (probe_id in names(plot_list)) {
  
  p <- plot_list[[probe_id]]
  
  # Sanitise the filename in case probe_id has characters like "/"
  safe_name <- gsub("[^A-Za-z0-9_\\-]", "_", probe_id)
  file_out  <- file.path(output_dir, paste0(safe_name,"_chartDis", ".pdf"))
  
  ggsave(
    filename = file_out,
    plot     = p,
    device   = "pdf",
    dpi      = 300,
    width    = 20,   # adjust if you want larger/smaller files
    height   = 6,
    units    = "in"
  )
}

cat("Saved", length(plot_list), "plots to", output_dir, "\n")

```

## Heatmap
```{r, Heatmaps, message=FALSE, warning=FALSE, fig.width = 20, fig.height = 25}

# ------------------------------------------------------------------
# 1)  Parameters
# ------------------------------------------------------------------
cells_per_group <- 3000    # max cells to keep *per cluster / sample* for heatmaps

# ---- by CLUSTER ---------------------------------------------------
Idents(UTSW_filtered) <- "seurat_clusters"

UTSW_ds_cluster <- subset(
  UTSW_filtered,
  downsample = cells_per_group   # Seurat takes ≤ N cells *per identity*
)

# ---- by SAMPLE ----------------------------------------------------
Idents(UTSW_filtered) <- "sample"

UTSW_ds_sample <- subset(
  UTSW_filtered,
  downsample = cells_per_group
)

# ------------------------------------------------------------------
# 2)  Find marker genes (same settings as before)
# ------------------------------------------------------------------

# --- cluster markers
Idents(UTSW_ds_cluster) <- "seurat_clusters"
cluster_markers <- FindAllMarkers(
  UTSW_ds_cluster,
  assay           = "ProbeCollapseAssay",
  only.pos        = TRUE,
  min.pct         = 0.10,
  logfc.threshold = 0.25
)

top_cluster_genes <- cluster_markers %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 10) %>%
  pull(gene) %>%
  unique()

# --- sample markers
Idents(UTSW_ds_sample) <- "sample"
sample_markers <- FindAllMarkers(
  UTSW_ds_sample,
  assay           = "ProbeCollapseAssay",
  only.pos        = TRUE,
  min.pct         = 0.10,
  logfc.threshold = 0.25
)

top_sample_genes <- sample_markers %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 10) %>%
  pull(gene) %>%
  unique()

# ------------------------------------------------------------------
# 3)  (Re)scale the genes to plot – still z-scores
# ------------------------------------------------------------------
heat_feats <- unique(c(top_cluster_genes, top_sample_genes))

UTSW_ds_cluster <- ScaleData(
  UTSW_ds_cluster,
  assay    = "ProbeCollapseAssay",
  features = heat_feats,
  verbose  = FALSE
)

UTSW_ds_sample <- ScaleData(
  UTSW_ds_sample,
  assay    = "ProbeCollapseAssay",
  features = heat_feats,
  verbose  = FALSE
)

# ------------------------------------------------------------------
# 4)  Draw heat-maps (now with far fewer columns → displays correctly)
# ------------------------------------------------------------------
heat_cluster <- DoHeatmap(
  UTSW_ds_cluster,
  features = top_cluster_genes,
  group.by = "seurat_clusters",
  slot     = "scale.data"
) +
  ggtitle("Top DE Genes per Cluster (down-sampled, z-scores)") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

heat_sample <- DoHeatmap(
  UTSW_ds_sample,
  features = top_sample_genes,
  group.by = "sample",
  slot     = "scale.data"
) +
  ggtitle("Top DE Genes per Sample (down-sampled, z-scores)") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

# ------------------------------------------------------------------
# 5)  Display
# ------------------------------------------------------------------

heat_cluster <- heat_cluster +
  theme(
    axis.text.y = element_text(
      size   = 6,              # smaller font
      margin = margin(r = 2)   # 2-pt right margin → label–tile gap
    )
)

heat_sample  <- heat_sample +
  theme(
    axis.text.y = element_text(size = 6, margin = margin(r = 2))
)

heat_cluster
heat_sample


## save figure
# save as PDF (vector) ──────────────
ggsave(
  filename = "figures/heat_cluster.pdf",
  plot     = heat_cluster,      # the object you created
  width    = 20,                # ← 20 inches wide
  height   = 25,                 # ← 25 inches tall
  units    = "in",              # "in", "cm", or "mm"
  device   = cairo_pdf          # good font embedding on most systems
)

ggsave(
  filename = "figures/heat_sample.pdf",
  plot     = heat_sample,      # the object you created
  width    = 20,                # ← 20 inches wide
  height   = 25,                 # ← 25 inches tall
  units    = "in",              # "in", "cm", or "mm"
  device   = cairo_pdf          # good font embedding on most systems
)

```

## Volcano Plots
```{r, volcano plots}

# ------------------------------------------------------------
# 0)  Parameters you might tweak
# ------------------------------------------------------------
logfc_cutoff <- 0.6   # |log2FC| threshold for "big" effect
padj_cutoff  <- 0.05  # FDR threshold
label_top    <- 10    # annotate this many most-sig genes per plot
save_plots   <- TRUE  # set FALSE if you don’t want files on disk

# ------------------------------------------------------------
# 1)  Prepare Seurat object and helper pkgs
# ------------------------------------------------------------
DefaultAssay(UTSW_filtered) <- "ProbeCollapseAssay"
Idents(UTSW_filtered)       <- "sample"

library(Seurat)
library(dplyr)
library(ggplot2)
library(ggrepel)    # nicer text labels
library(patchwork)  # just in case
library(purrr)      # iterate/keep tidy (map())

# ------------------------------------------------------------
# 2)  Iterate over samples → DE test vs. all others
# ------------------------------------------------------------
sample_levels <- levels(UTSW_filtered$sample)

volcano_list <- map(sample_levels, function(smpl) {
  
  message("Finding markers for ", smpl, " vs rest …")
  
  deg <- FindMarkers(
    object      = UTSW_filtered,
    ident.1     = smpl,
    group.by    = "sample",
    assay       = "ProbeCollapseAssay",
    only.pos    = FALSE
  ) %>%
    rownames_to_column("gene") %>%
    mutate(
      sign      = case_when(
                    p_val_adj < padj_cutoff & avg_log2FC >  logfc_cutoff ~ "Up",
                    p_val_adj < padj_cutoff & avg_log2FC < -logfc_cutoff ~ "Down",
                    TRUE                                                     ~ "NS"
                  ),
      negLog10P = -log10(p_val_adj + 1e-300)   # +tiny to avoid Inf
    )
  
  ## Top genes to label (most significant)
  lab_genes <- deg %>%
    filter(sign != "NS") %>%
    arrange(p_val_adj) %>%
    slice_head(n = label_top) %>%
    pull(gene)
  
  ## Volcano plot
  p <- ggplot(deg, aes(x = avg_log2FC, y = negLog10P)) +
    geom_point(aes(color = sign), alpha = 0.6, size = 1) +
    scale_color_manual(values = c(Up = "#D62728", Down = "#1F77B4", NS = "grey70")) +
    geom_vline(xintercept =  c(-logfc_cutoff, logfc_cutoff), linetype = 2) +
    geom_hline(yintercept = -log10(padj_cutoff), linetype = 2) +
    geom_text_repel(
      data  = filter(deg, gene %in% lab_genes),
      aes(label = gene),
      max.overlaps = Inf,
      size  = 3
    ) +
    labs(
      title    = paste("Volcano:", smpl, "vs all other samples"),
      x        = "avg log2 Fold-Change",
      y        = "-log10(adj. P value)",
      color    = ""
    ) +
    theme_classic() +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  
  p
})

names(volcano_list) <- sample_levels

# ------------------------------------------------------------
# 3)  Optional: save each plot to disk
# ------------------------------------------------------------
if (save_plots) {
  dir.create("volcano_plots", showWarnings = FALSE)
  
  walk2(volcano_list, names(volcano_list), function(p, nm) {
    ggsave(
      filename = file.path("volcano_plots", paste0("volcano_reg_", nm, ".pdf")),
      plot     = p,
      width    = 6, height = 5, units = "in", dpi = 300
    )
  })
  message("Saved ", length(volcano_list), " volcano plots to volcano_plots/")
}

# ------------------------------------------------------------
# 4)  Preview: show first plot inline
# ------------------------------------------------------------
volcano_list

```

## Volcano Plots Genes of Interest
```{r, GOI Volcano Plots}

# ──────────────────────────────────────────────────────────────
# Volcano plots (genes of interest) + SAVE as vector files
# ──────────────────────────────────────────────────────────────
# prerequisites: UTSW_filtered, genes_probes, logfc_cutoff, padj_cutoff
# returns: named list `volcano_list` and writes PDF/SVG to disk
# ----------------------------------------------------------------
DefaultAssay(UTSW_filtered) <- "ProbeCollapseAssay"
Idents(UTSW_filtered)       <- "sample"

library(Seurat)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(purrr)
library(svglite)   # for SVG output

sample_levels <- levels(UTSW_filtered$sample)

volcano_list <- map(sample_levels, function(smpl) {

  ## 1) Differential expression
  deg <- FindMarkers(
           UTSW_filtered,
           ident.1  = smpl,
           group.by = "sample",
           assay    = "ProbeCollapseAssay",
           only.pos = FALSE
         ) %>%
         rownames_to_column("gene") %>%
         mutate(
           sign      = case_when(
                          p_val_adj < padj_cutoff & avg_log2FC >  logfc_cutoff ~ "Up",
                          p_val_adj < padj_cutoff & avg_log2FC < -logfc_cutoff ~ "Down",
                          TRUE                                                      ~ "NS"
                        ),
           negLog10P = -log10(p_val_adj + 1e-300)
         )

  ## 2) Label genes of interest that are significant
  label_df <- deg %>%
                filter(
                  gene %in% genes_probes,
                  sign != "NS"
                )

  ## 3) Plot
  ggplot(deg, aes(avg_log2FC, negLog10P)) +
    geom_point(aes(color = sign), alpha = 0.6, size = 1) +
    scale_color_manual(values = c(Up = "#D62728", Down = "#1F77B4", NS = "grey70")) +
    geom_vline(xintercept = c(-logfc_cutoff, logfc_cutoff), linetype = 2) +
    geom_hline(yintercept = -log10(padj_cutoff),            linetype = 2) +
    geom_text_repel(
      data  = label_df,
      aes(label = gene),
      max.overlaps = Inf,
      size = 3
    ) +
    labs(
      title = paste("Volcano – Genes of Interest:", smpl, "vs rest"),
      x     = "avg log2 Fold-Change",
      y     = "-log10(adj. P value)",
      color = ""
    ) +
    theme_classic() +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
})

names(volcano_list) <- sample_levels

# ──────────────────────────────────────────────────────────────
# Save each volcano plot as PDF
# ──────────────────────────────────────────────────────────────
out_dir <- "volcano_GOI_plots"
dir.create(out_dir, showWarnings = FALSE)

plot_w <- 6   # inches
plot_h <- 5

walk2(volcano_list, names(volcano_list), function(p, nm) {
  safe_nm <- gsub("[^A-Za-z0-9_\\-]", "_", nm)
  
  ggsave(
    filename = file.path(out_dir, paste0("volcano_GOI_", safe_nm, ".pdf")),
    plot     = p,
    device   = cairo_pdf,
    width    = plot_w, height = plot_h, units = "in"
  )
})

message("Saved ", length(volcano_list) * 2,
        " vector files to ", normalizePath(out_dir))
```

## Full‐dataset DGE and export
```{r full_DGE_export, message=FALSE, warning=FALSE}
# ensure you’re using your collapsed probe assay
DefaultAssay(UTSW_filtered) <- "ProbeCollapseAssay"

# make output folder if needed
if (!dir.exists("results")) dir.create("results")

# 1) Cluster‐level DGE on full data
Idents(UTSW_filtered) <- "seurat_clusters"
cluster_markers_full <- FindAllMarkers(
  object          = UTSW_filtered,
  assay           = "ProbeCollapseAssay"
)
write.csv(
  cluster_markers_full,
  file      = "results/cluster_DGE_full.csv",
  row.names = FALSE
)

# 2) Sample‐level DGE on full data
Idents(UTSW_filtered) <- "sample"
sample_markers_full <- FindAllMarkers(
  object          = UTSW_filtered,
  assay           = "ProbeCollapseAssay"
)
write.csv(
  sample_markers_full,
  file      = "results/sample_DGE_full.csv",
  row.names = FALSE
)

message("Wrote cluster_DGE_full.csv and sample_DGE_full.csv to results/") 
```

## Export as loupe file
```{r eval=FALSE, include=FALSE}

library(loupeR)
output_dir = '/users/c/o/cookesf/008_R-Studio_Projects/2025-04-15_UTSW_Figures'
# Generate Loupe Browser file using create_loupe_from_seurat
create_loupe_from_seurat(
  obj = UTSW_filtered,
  output_dir = output_dir,
  output_name = "UTSW_Adam_Figures_ProbeCollapsed_5_Removed_filtered",
  dedup_clusters = FALSE,
  feature_ids = NULL,
  executable_path = NULL,
  force = FALSE
)

```

